---
title: "Exact Statistics and Semi-Parametric Tests for Small Network Data"
author:
  - \textbf{George G. Vega Yon, MS}$^\star$
  - Andrew Slaughter, PhD$^\dagger$
  - Kayla de la Haye, PhD$^\star$
date: "Sunbelt 2019, Montreal \\linebreak[4]June 20, 2019"
# institute: "\\includegraphics[width=.15\\linewidth]{usc.pdf} \\hspace{.5cm} \\includegraphics[width=.15\\linewidth]{sunbelt-logo.pdf}"
institute: "\\begin{minipage}[c]{.3\\linewidth}\\centering $^\\star$University of Southern California\\linebreak[4]Department of Preventive Medicine\\end{minipage}\\begin{minipage}[c]{.3\\linewidth}\\centering $^\\dagger$U.S. Army Research Institute for the Behavioral and Social Sciences\\end{minipage}"
output:
  beamer_presentation:
    slide_level: 2 # revealjs::revealjs_presentation
    highlight: espresso
    latex_engine: xelatex
    includes:
      in_header: notation-def.tex
aspectratio: 169
# handout: false
fontsize: 10pt
bibliography: bibliography.bib
colortheme: beaver
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(smallsize = function(before, options, envir) {
    if (before) {
        "\\footnotesize\n\n"
    } else {
        "\n\\normalsize\n\n"
    }
})
knitr::opts_chunk$set(echo = TRUE, smallsize=TRUE)
```

## Acknowledgements

\scriptsize

\begincols
\begincol{.2\linewidth}

\includegraphics[width=.7\linewidth]{fig/ARO_logo.png}

\endcol

\begincol{.79\linewidth}
This material is based upon work support by, or in part by, the U.S. Army Research
Laboratory and the U.S. Army Research Office under grant number W911NF-15-1-0577
\endcol
\endcols

\begincols
\begincol{.79\linewidth}

Computation for the work described in this paper was supported by the University
of Southern California’s Center for High-Performance Computing (hpc.usc.edu).
\endcol

\begincol{.2\linewidth}
\includegraphics[width = .7\linewidth]{fig/usc.pdf}
\endcol
\endcols

The views expressed in this presentation are those of the authors, and do not represent the official policy or positions of the Department of the Army, the DoD, or the U.S. Government.

\begin{figure}
\centering
\includegraphics[width = .85\linewidth]{fig/muriteams.png}
\end{figure}

## Context: Social abilities and team performance

Two research questions\pause

\begin{centering}

\large How do \uscred{social abilities} impact \uscred{network structure}? \pause

\color<4->{gray}{How does \only<3>{\uscred{collective intelligence}}\only<4->{collective intelligence} affect team (network) \only<3>{\uscred{performance}}\only<4->{performance}?}\pause
\normalsize

\end{centering}

---

To answer this question, we have the following experimental data: \pause

- 42 mixed-gender teams,\pause

- Which completed 1 hour of group tasks (MIT's IQ test for teams)\pause

- Individual survey capturing information regarding socio-demographics **and**:\pause
    
    - \uscred{Social Intelligence}: Social Perception (measured by RME), Social Accomodation, Social Gregariousness, and Social Awareness \pause
    
    - \uscred{Social Networks}: Advice Seeking, Leadership, Influence (among others).

## Context  (cont'd)

\begin{figure}
\centering
\includegraphics[width = .6\linewidth]{fig/plot-graph-4-1.pdf}
\end{figure}

We can do a lot of simple statistics: density, \% of \textit{\color{gray}[blank]}, etc. but...\pause{} \uscred{\large how can we go beyond that?}

## Exponential random graph models

\def\fig1width{.35\linewidth}
\begin{figure}
\centering
\begin{tabular}{m{.2\linewidth}<\centering m{.4\linewidth}<\raggedright}
\toprule Representation & Description  \\ \midrule
\includegraphics[width=\fig1width]{terms/mutual.pdf} & Mutual Ties (Reciprocity)\linebreak[4]$\sum_{i\neq j}y_{ij}y_{ji}$  \\
\includegraphics[width=\fig1width]{terms/ttriad.pdf} & Transitive Triad (Balance)\linebreak[4]$\sum_{i\neq j\neq k}y_{ij}y_{jk}y_{ik}$  \\
\includegraphics[width=\fig1width]{terms/homophily.pdf} & Homophily\linebreak[4]$\sum_{i\neq j}y_{ij}\mathbf{1}\left(x_i=x_j\right)$ \\
\includegraphics[width=\fig1width]{terms/nodeicov.pdf} & Covariate Effect for Incoming Ties\linebreak[4]$\sum_{i\neq j}y_{ij}x_j$ \\
\includegraphics[width=\fig1width]{terms/fourcycle.pdf} & Four Cycle\linebreak[4]$\sum_{i\neq j \neq k \neq l}y_{ij}y_{jk}y_{kl}y_{li}$  \\
\bottomrule
\end{tabular}
\end{figure}

ERGMs can do the job, but the only problem is...\pause{} have you tried estimating ERGMs on small networks?

## Exponential random graph models for small networks

A lot of\pause

-  Playing with the MCMC control parameters to obtain sensible statistics, or\pause

-  Sometimes we also go for using a single big (very sparse) graph\pause
    
    -  Block diagonal matrix 
    
    -  Constrain the sampling space puting structural zeros\pause
    
This fails too often (smaller networks = higher chance of model degeneracy).

## Revising model degeneracy

Following @Handcock2003, the key question is: Where do the sufficient statistics live?

\begin{figure}
\centering
\includegraphics[height=.6\textheight]{handcock2003-figure1.pdf}
\end{figure}

\pause

-  In the interior: \uscred{Good}, we (possibly) get nice estimates in both MC-MLE and MLE \pause

-  Not in the interior: \uscred{We are in trouble}, we mostly get degenerate estimates (more with MC-MLE, but still with MLE) 

## ERGMs for small networks

*   Calculating the likelihood function for a directed graph means (at some point)
    enumerating \uscred{$2^{n(n-1)}$ terms}.
    
    $$
    \Prcond{\Graph = \graph}{\params, \Indepvar} = \frac{%
    		\exp{\transpose{\params}\sufstats{\graph, \Indepvar}}%	
    	}{
    		\color{USCCardinal}\sum_{\graph'\in\GRAPH} \exp{\transpose{\params}\sufstats{\graph', \Indepvar}}\color{black}
    	}
    $$
    
    \pause

*   So, if $n = 6$, then we have approx 1,000,000,000 terms \includegraphics[width=.04\linewidth]{scared.pdf}.\pause 

*   This has lead the field to aim for (very neat) simulation based methods\pause

*   But, if our small networks have (at most) 6 nodes... \pause

\large \uscred{We can go back to the \textbf{good-old-fashion} MLE}\normalsize \raggedleft

---


Keeping $n\leq 6$ we can\pause

-  Compute the likelihood function exactly, and hence use ``simple'' optimization to get MLEs.\pause

-  Obtain more \uscred{accurate} estimates \uscred{faster}\pause{} (in most cases). \pause

-  Since (usually) small networks come in many...\pause obtain pooled estimates. Which helps with \uscred{power} \textit{and} \uscred{degeneracy})\pause

-  etc.\pause

This and more has been implemented in the `ergmito` (\includegraphics[width=.1\linewidth]{fig/lifecycle-experimental-orange.pdf}) R package (available at https://github.com/muriteams/ergmito) \pause

(built on top of Statnet's amazing `ergm` @hunter2008; @Handcock2018 R package)

---

Sidetrack...

\begin{minipage}[c]{1\linewidth}
\large \textbf{ito, ita}: From the latin -\textit{\=ittus}. suffix in Spanish used to denote small or affection. e.g.:

\hspace{.5cm} \textit{¡Qué lindo ese perr\textcolor{USCCardinal}{\textbf{ito}}!} / \textit{What a beautiful little dog!}

\hspace{.5cm} \textit{¿Me darías una tac\textcolor{USCCardinal}{\textbf{ita}} de azúcar?} / \textit{Would you give me a small cup of sugar?}
\normalsize
\end{minipage}\pause

\alert{Special thanks to George Barnett who proposed the name during the 2018 NASN!}


## Quick example

```{r loading-fivenets, include=FALSE}
library(ergmito)
data(fivenets, package = "ergmito")
```

Suppose that we have 5 `network`s (as in the R package `network`)

```{r plotfivenets, warning=FALSE, message=FALSE, echo=FALSE, fig.width=6, fig.height=3, out.width='.5\\linewidth', fig.align='center', cache=TRUE}
library(sna)
library(network)
op <- par(mfrow = c(2, 3), mai=rep(0, 4), oma = rep(0, 4))
USCCARDINAL <- rgb(153, 0, 0, maxColorValue = 255)
ans <- lapply(fivenets, function(f) {
  gplot(
    f,
    vertex.cex = 2,
    vertex.col = c("white", USCCARDINAL)[
      get.vertex.attribute(f, "female") + 1
    ]
    )
  })
plot.new()
plot.window(xlim = c(0, 1), ylim = c(0, 1))
legend("center", fill = c("white", USCCARDINAL), legend = c("Male", "Female"), cex=1, bty="n")
par(op)
```

\begin{minipage}[c]{.6\linewidth}
And we would like to fit a model using the edgecount and number of gender-homophilic ties.\pause
\end{minipage}

\raggedleft \large How can we do it?

## `ergmito` example (cont'd)

The same as you would do with the `ergm` package\pause{}

```{r fit-fivenets, cache=TRUE}
model1 <- ergmito(fivenets ~ edges + nodematch("female"))
```

\pause

```{r fit-fivenets-print, echo=TRUE, cache=TRUE}
summary(model1) # 
```

\pause

Go to \url{https://github.com/muriteams/ergmito} for more on this R package.
<!-- --- -->

<!-- ```{r gof-fivenets-print, cache=TRUE, out.width=".7\\linewidth", fig.align='center', width=8, height=6} -->
<!-- gof1 <- gof_ergmito(model1) -->
<!-- plot(gof1) -->
<!-- ``` -->

## How many networks?

-  Thinking about power and unbiasedness, we did a simulation study

-  Simulated 100,000 samples of networks using the following steps:

    1.  Draw parameters for _edges_ and _mutual_ from a uniform(-3, 3).
    
    2.  Draw group sizes $n_1\sim\mbox{Poisson}(10), n_2\sim\mbox{Poisson}(10), n_3\sim\mbox{Poisson}(10)$, networks of size 3, 4, and 5 respectively.

    3.  Using 1. and 2., simulate networks using ERGM
    
-  We looked at empirical bias (sanity check), and power


## How many networks? Bias

\begin{figure}
\centering
\includegraphics[width=.65\linewidth]{fig/bias-02-various-sizes-3-5.pdf}
\end{figure}

## How many networks? Power

\begin{figure}
\centering
\includegraphics[width=.65\linewidth]{fig/power-02-various-sizes-3-5.pdf}
\end{figure} \pause{}

What about a real data set?

## Preliminary results

From our sample of 42 small networks:

```{r muri-pre, cache=TRUE, echo=FALSE, results='asis'}
ans       <- readRDS("model/ergmito_forward_selection_part4.rds")
library(ergmito)

names(ans) <- gsub("ans_", "", names(ans))
texreg::texreg(
  ans, booktabs = TRUE, fontsize = "tiny",
  use.packages = FALSE, caption = "Selected models for each one of the studied networks. Results presented here correspond to a forward selection process."
  )
```

----

```{r muri-gof, cahe=TRUE, echo=FALSE, fig.align='center', out.width=".8\\linewidth"}
load("model/ergmito_forward_selection_part4_gof.rda")
plot(gof_advice, main = "GOF Advice seeking network")
```


## Context: Social abilities and team performance

Two research questions\pause

\begin{centering}

\large \sout{How do \uscred{social abilities} impact \uscred{network structure}?} 

How does \uscred{collective intelligence} affect team (network) \uscred{performance}?
\normalsize

\end{centering}

## Networks and team performance

Suppose we have the following:

-  Data on structure, nodes, and an outcome: $\left(\graph, \indepvar, \depvar\right)$\pause

-  In general, we are interested on assessing the following: $\graph\perp\depvar$ \pause

-  Ways to solve this: parametrically (e.g. GLMs) and non-parametrically (permutation tests):\pause

- For parametric approaches: Sample size?\pause
    
- Non-parametrically: Control for confounders ($\indepvar \to\depvar, \indepvar\to\graph$)?\pause

Perhaps ERGMs can help us here (to generate null distributions)

<!-- ## Networks and team performance -->

<!-- Imagine that: -->

<!-- - $\indepvar_i$: Individual level CI -->

<!-- - $\graph \sim \mbox{ERGM}\left(\mbox{indegree of CI, edgecount}\right)$ -->

<!-- - $\mbox{E}\left\{\depvar \right\} = $ -->

---

\footnotesize

\def\svgwidth{.8\linewidth}
\begin{figure}
\centering
\input{fig/struct-test.pdf_tex}
\end{figure}\pause

\normalsize

<!-- \begin{minipage}[c]{.7\linewidth} -->
<!-- We can use the distribution of the sequence $\{t_1, \dots, t_B\}$ as null to compare against $t_0$\pause -->
<!-- \end{minipage} -->

We are still working (thinking) about this...



## Discussion

*   ERGMItos... This is not new.\pause{} What's new is the set of tools to apply it\pause

*   Taking this approach we can improve our estimates (power) and help with
    degeneracy\pause

*   The tool is working\pause{} (according to the simulation study...)\pause

*   Need to conduct more simulations using _nodal_ attributes and compare with ERGM block diagonal models.\pause

*   What about goodness-of-fit? Still need to better think about it
    

## Discussion (contd')

*  The simplicity of the estimation procedure allows us to think of:\pause

    -  Separable Temporal ERGMitos, a.k.a. TERGMitos \pause

    -  Mixture models and Bayesian inference (if you are into that kind of stuff)\pause

    -  More flexible formulas (e.g. interactions between terms and graph-level attributes) \pause{}

    -  Better odds ratios (not simply exponentiating the coefficients)\pause

    -  Simulation based methods (small size $\implies$ sampling from in-memory data, and exact tests)\pause
    
    -  Cross-validation/model selection in ERGMs (thank you, Nolan \includegraphics[width=.04\linewidth]{vader.png}!)\pause{}

*  Still thinking about how to test for association between network structure and group outcome

## Thanks!

\begin{centering}

We thank members of our MURI research team, USC's Center for Applied Network Analysis, Garry Robins, Carter Butts, Johan Koskinen, Noshir Contractor, and attendees of the NASN 2018 conference for their comments.

\includegraphics[width=.2\linewidth]{logo.png}

\def\and2{\hskip 1em}

\large \textbf{George G. Vega Yon, MS} \and2 Andrew Slaughter, PhD \and2 Kayla de la Haye, PhD

\href{mailto:vegayon@usc.edu}{vegayon@usc.edu} 

\href{https://ggvy.cl}{https://ggvy.cl} 

\includegraphics[width=.02\linewidth]{github.png}\href{https://github.com/gvegayon}{gvegayon}
\includegraphics[width=.02\linewidth]{twitter.png}\href{https://twitter.com/gvegayon}{gvegayon}

\end{centering}


## References
